---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
The GitHub repository contains the dataset, so unzip it, read it (and cast the date column as a date object so it can be easily plotted).
```{r}
unzip("activity.zip")
activity <- read.csv("activity.csv")

#library(tidyr)
#library(lubridate)
#activity <- mutate(activity, date = factor(ymd(date)))
```
## What is mean total number of steps taken per day?
Summarize the data so there is a total sum of steps for each day. Then produce the histogram showing the distribution of total daily steps.
```{r}
library(ggplot2)
suppressMessages(library(dplyr))

daily.activity <- activity %>%
    group_by(date) %>%
    summarize(steps = sum(steps, na.rm = TRUE))

g1 <- ggplot(daily.activity, aes(x = steps)) +
    geom_histogram(binwidth = 2000, fill = "steelblue") +
    labs(title = "Histogram of total steps taken each day")
g1
```

Calculate the mean and the median of the total number of steps per day. Store this as a vector so that we can reuse it for a later comparison.
```{r}
original.ave <- c(mean = round(mean(daily.activity$steps)),
                  median = round(median(daily.activity$steps)))
original.ave
```

## What is the average daily activity pattern?
Summarize the average number of steps for each 5 minute interval within a day, and plot it.
```{r}
ia <- activity %>%
    group_by(interval) %>%
    summarize(steps = mean(steps, na.rm = TRUE))

g2 <- ggplot(ia, aes(x = interval, y = steps)) +
    geom_line(colour = "steelblue") +
    labs(title = "Average steps for each 5 minute interval, averaged across all days")
g2
```

Calculate the interval containing the maximum number of steps, and average across all days. Using the mode would be the most appropriate method here. We'll calculate the mean and median as well, for comparison.
```{r}
max.activity <- activity %>%
    group_by(date) %>%
    # which.max() causes an error if all values for a day are NA, because it 
    # returns an integer instead of a vector of size 1. 
    # Adding "0" to the end of the steps vector prevents this, and returns NA
    # when we attempt to lookup that index in the interval vector.
    summarize(interval = interval[which.max(c(steps, 0))]) 

suppressMessages(library(modeest))
c(mode = mlv(max.activity$interval, na.rm = TRUE)$M,
  median = median(max.activity$interval, na.rm = TRUE),
  mean = round(mean(max.activity$interval, na.rm = TRUE)))
```

The mode of these values is `r mlv(max.activity$interval, na.rm = TRUE)$M`.

## Imputing missing values
Calculate the number of missing values in the dataset.
```{r}
sum(is.na(activity$steps))
```

To impute the missing values we will replace them with the mean value for that 5 minute interval. This can be looked up from `ia` generated in the second part of this analysis. The mean and median values can be compared with the values calculated in the first part of this analysis
```{r}

act2 <- activity
act2$na <- is.na(act2$steps)
act2$steps <- replace(act2$steps,
                      act2$na,
                      # Reuse ia from the second part of this analysis.
                      round(ia$steps[ia$interval == act2$interval[act2$na]]))

daily.act2 <- act2 %>%
    group_by(date) %>%
    summarize(steps = sum(steps, na.rm = TRUE))

imputed.ave <- c(mean = round(mean(daily.act2$steps)),
                 median = round(median(daily.act2$steps)))
data.frame(original.ave, imputed.ave)
```

This shows that the mean and median values have increased. This is because the NA values originally contributed zero to the daily total. In the imputed data, the mean value for that interval is used instead, increasing the totals. 

```{r}
g3 <- ggplot(daily.act2, aes(x = steps)) +
    geom_histogram(binwidth = 2000, fill = "chocolate") +
    labs(title = "Histogram of total steps taken each day with NAs imputed")
g3
```

The histogram above shows that the distribution of total steps hasn't changed a lot after imputing the NA values. A single day has moved from the bin including zero to the bin including the mean. Presumably this day had most or all values as zero, so the majority of values in this day were modified.


## Are there differences in activity patterns between weekdays and weekends?
